A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-15 paper/paper.tex
Minor
Add admissible gray boxes

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper. We find de Bruijn
representation advantageous for specification since it leaves very
little ambiguity about the variable freeness side
conditions, making our proof more easily reproducible. Furthermore, as we discuss in
Section~\ref{sec:automation}, de Bruijn representation is much more
amenable to automated reasoning in proof assistants.
We adopt from \citet{autosubst} the notations for simultaneous
renaming, substitution, and other auxiliary definitions related to our
syntax, summarized in Figure~\ref{fig:auxdef}. We omit most of the
definitions of renaming and substitution and only show the definition
of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \mathit{IdentityRen} \\
        [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm i]] & := & [[var i]] \\ \\
        \mathit{ConsRen ([[xi .: n]])} \\
        ([[xi .: n]])([[0]]) & := & [[n]] \\
        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n i]] & := & [[n]] + [[i]]
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{UpRen} \\
        [[( up xi ) 0]] & := & [[0]] \\
        [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        \mathit{UpTm} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Renaming ([[a < xi >]])} \\
        [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        \ldots \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT, where the judgmental equality typically
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. Furthermore, it is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types). % and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.

@@

\section{Logical Relation}
\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}). The
logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures:
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.

@@

The following lemma makes the statement $[[G |= A : Set i]]$ easier to
work with.
\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}]
  \label{lemma:setinv}
  The following two statements are equivalent:
  \begin{itemize}
  \item $[[G |= A : Set i]]$
  \item For all $[[rho]]$, if $[[rho |= G]]$, then there exists
    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The forward direction is immediate by
  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
  and show that $[[G |= A : Set i]]$ given the second bullet.

@@

In this section, we sketch out how the logical relation from
Section~\ref{sec:logreldep} can be extended show the existence of
$\beta\eta$ normal forms for (open and closed) well-typed terms. We
first extend parallel reduction to include $\eta$ reduction for
functions.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
equational theory since coherence, the untyped relation used for type
conversion, is built on top of parallel reduction. We can recover the
same confluence result about parallel reduction using the standard
techniques from \citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though specifically for the de Bruijn
representation, we need the following anti-renaming lemma about
parallel reduction.
\begin{lemma}[Par anti-renaming]
  \label{lemma:parantirenaming} If $[[a { xi } => b0]]$, then there
exists some $[[b]]$ such that $[[b { xi } = b0]]$ and $[[a => b]]$.
\end{lemma}

@@

Interestingly, all the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} up to the
fundamental lemma can be proven in the exact same order, where the new
cases related due to \rref{I-Ne} and the augmentation of neutral terms
to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
Lemma~\ref{lemma:parnenf}.

@@

\begin{definition}[CR]
  Let $[[S]]$ be a set of lambda terms. We say that
  \begin{itemize}
  \item $[[S]] \in CR_1$ if $[[wne a implies a in S]]$
  \item $[[S]] \in CR_2$ if $[[a in S implies wn a]]$
  \item $[[S]] \in CR$ if $[[S]] \in CR_1$ and $[[S]] \in CR_2$
  \end{itemize}
\end{definition}

@@

\begin{lemma}[$CR$ for terms (Rec)]
  \label{lemma:crelrec}
  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  After unfolding the definition of $[[InterpR i A S]]$, trivial by
  Lemma~\ref{lemma:cr2ty}, \ref{lemma:crel} and \ref{lemma:cr1ty}.
\end{proof}

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal form]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]](i) = i$, which injects
  natural numbers as term variables. This is
  allowed since Lemma~\ref{lemma:crelrec} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho} = A]]$ and $[[a {rho} = a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}
From Corollary~\ref{corollary:exbetanf}, we can show that there exists
$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
and also strictly decreases the size of the term.

@@

Finally, we want to point out that
Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
\ref{lemma:crelrec}, often referred to as adequacy of the logical
relation, are in fact required in the normalization proof for simply
typed languages~\citep{abel2019poplmark}. In the simply typed
scenario, adequacy needs to be proven before the fundamental theorem
so we can handle rules such as \rref{T-If} where the scrutinee is a
neutral term. In a simply typed setting, Lemma~\ref{lemma:extwn} is
used in the exact same way to show that lambda terms are themselves
normalizing as we have done in Lemma~\ref{lemma:crel}.

@@

However, we choose instead an axiom-free workaround and define
\rref{I-Pi} as follows in our Coq mechanization.
\begin{center}
  \drule[]{I-PiCoq}
\end{center}
It should be easy to verify that the preconditions of
\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
equivalent. After establishing Lemma~{lemma:logreldeter}, it is
possible to further show that the conclusions of the rules are
equivalent, too.

@@

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We hope features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts can be implemented in a similarly streamlined
fashion and we will leave those as part of our future work.

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage{tcolorbox}
\usepackage[para]{footmisc}
\definecolor{lightgray}{gray}{0.75}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper. We find de Bruijn
representation advantageous for specification since it leaves very
little ambiguity about the variable freeness side
conditions, making our proof more easily reproducible. Furthermore, as we discuss in
Section~\ref{sec:automation}, de Bruijn representation is much more
amenable to automated reasoning in proof assistants.
We adopt from \citet{autosubst} the notations for simultaneous
renaming, substitution, and other auxiliary definitions related to our
syntax, summarized in Figure~\ref{fig:auxdef}. We omit most of the
definitions of renaming and substitution and only show the definition
of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \mathit{IdentityRen} \\
        [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm i]] & := & [[var i]] \\ \\
        \mathit{ConsRen ([[xi .: n]])} \\
        ([[xi .: n]])([[0]]) & := & [[n]] \\
        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n i]] & := & [[n]] + [[i]]
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{UpRen} \\
        [[( up xi ) 0]] & := & [[0]] \\
        [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        \mathit{Renaming ([[a < xi >]])} \\
        [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        \ldots \\ \\
        \mathit{UpTm} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
In \rref{T-Var}, $[[G i]]$ is the partial function defined
with the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
i ]] := [[G i]] $. The precondition $[[i < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
i]]$ has a defined output.
\Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. It is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types). % and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.

@@

\section{Logical Relation}
\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}).
Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
The logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.

@@

The following lemma makes the statement $[[G |= A : Set i]]$ easier to
work with.
\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}]
  \label{lemma:setinv}
  The following two statements are equivalent:
  \begin{itemize}
  \item $[[G |= A : Set i]]$
  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The forward direction is immediate by
  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
  and show that $[[G |= A : Set i]]$ given the second bullet.

@@

In this section, we sketch out how the logical relation from
Section~\ref{sec:logreldep} can be extended show the existence of
$\beta\eta$ normal forms for (open and closed) well-typed terms. We
first extend parallel reduction to include $\eta$ reduction for
functions.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
equational theory since coherence, the untyped relation used for type
conversion, is built on top of parallel reduction. We can recover the
same confluence result about parallel reduction using the standard
techniques from \citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though specifically for the de Bruijn
representation, we need the following anti-renaming lemma about
parallel reduction.
\begin{lemma}[Par anti-renaming]
  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

@@

Interestingly, all the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the
fundamental lemma can be proven in the exact same order, where the new
cases due to \rref{I-Ne} and the augmentation of neutral terms
to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
Lemma~\ref{lemma:parnenf}.

@@

\begin{definition}[CR]
  Let $[[S]]$ be a set of lambda terms. We say that
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
  \end{itemize}
\end{definition}

@@

\begin{lemma}[$CR$ for terms (Rec)]
  \label{lemma:crelrec}
  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  After unfolding the definition of $[[InterpR i A S]]$, trivial by
  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
\end{proof}

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal form]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
  natural numbers as term variables. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}
From Corollary~\ref{corollary:exbetanf}, we can show that there exists
$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
and also strictly decreases the size of the term.

@@

Finally, we want to point out that
Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
\ref{lemma:crelrec}, often referred to as adequacy of the logical
relation, are in fact required in the normalization proof for simply
typed languages~\citep{abel2019poplmark}. Similar to the simply typed
scenario, adequacy needs to be proven before the fundamental theorem
so we can handle rules such as \rref{T-If} where the scrutinee is a
neutral term. In \citet{abel2019poplmark}, a variant of
Lemma~\ref{lemma:extwn} is used in the exact same way to show that
lambda terms are themselves normalizing as we have done in
Lemma~\ref{lemma:crel}.

@@

However, we choose instead an axiom-free workaround and define
\rref{I-Pi} as follows in our Coq mechanization.
\begin{center}
  \drule[]{I-PiCoq}
\end{center}
It should be easy to verify that the preconditions of
\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
possible to further show that the conclusions of the rules are
equivalent, too.

@@

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We hope features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts can be implemented in a similarly streamlined
fashion and we will leave those extensions as part of our future work.
